---
title: "Multi-class Conformal Classification"
date: 12/12/2025
author: "Nonparametric statistics ay 2025/2026"
output:
  
  html_document: 
    df_print: paged
    toc: true
  pdf_document: default
  html_notebook: 
    df_print: paged
  word_document: default
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(rgl)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
knitr::knit_hooks$set(webgl = hook_webgl)
ggplot2::theme_set(ggplot2::theme_bw())
```

```{css, echo=FALSE}
.extracode {
background-color: lightblue;
}
```

## Loading necessary libraries

```{r message=FALSE, warning=FALSE}
library(dplyr) 
library(ggplot2)
library(knitr)
library(broom)
library(tidyr)
library(progress)
library(pbapply)
pboptions(type='none')
library(dbscan)
library(gridExtra)

library(stats)
library(randomForest)
```

\pagebreak


## Multi-class Conformal Classification
The following R implementation directly descends from the methods for conformal classification discussed in:

Romano, Y., Sesia, M., & Candes, E. (2020). Classification with valid and adaptive coverage. Advances in Neural Information Processing Systems, 33, 3581-3591.

[Classification with valid and adaptive coverage](https://proceedings.neurips.cc/paper/2020/file/244edd7e85dc81602b7615cd705545f5-Paper.pdf)

We present below a simplified version of multi-class conformal classification. In this simplified version, we work in split-conformal framework and we make a specific choice of non-conformity scores, namely the *homogeneous nonconformity scores*. They are called *homogeneous* because they are not designed to account for heteroscedasticity in the distribution of $Y \mid X$. Therefore, for classification settings with heteroscedasticity, you should rely on more refined definitions of nonconformity scores. A Python implementation of the conformal classification method, as well as useful extensions for alternative definitions of nonconformity scores, can be found at:

[arc GitHub repository](https://github.com/msesia/arc)

### Conformal classification with *homogeneous nonconformity scores*
Let $(X_i,Y_i)_{i=1}^{n+1}$, where $X_i \in \mathbb{R}^{p}$ is a matrix of features and $Y_i \in \{1,\dots,K \}$ is a label.

Given a significance level $\alpha$, we want to identify a set $\hat{C}_{1-\alpha} (X_{n+1})$ with marginal coverage at level $1 - \alpha$. That is,
\begin{align*}
    \mathbb{P}\left[ Y_{n+1} \in \hat{C}_{1-\alpha}(X_{n+1})\right] \geq 1-\alpha.
\end{align*}

**Split-conformal classification**

We split the data in training set $D^{\mathrm{train}}$ and calibration set $D^{\mathrm{cal}}$.

$D^{\mathrm{train}}$ is used to fit a (black-box) classification model $\hat{\pi}$:
\begin{align*}
    \hat{\pi}_k(x) = \hat{\mathbb{P}}\left[Y=k \mid X=x\right].
\end{align*}
Without loss of generality, assume that $\hat{\pi}$ is normalized; i.e., $\sum_{k=1}^{K} \hat{\pi}(x,k) = 1$.

Having pre-trained the model $\hat{\pi}$ on the data in $D^{\mathrm{train}}$, we use the calibration data in $D^{\mathrm{cal}}$ to compute the nonconformity scores $\hat{s}(X_i, k)$, for all $i \in [n]$.
In particular, we consider a specific definition of nonconformity scores, namely the ones originating from the definition of *homogeneous score function*.

For any label $k \in [K]$ and a vector of features $x$, the *homogeneous score function* with respect to label $k$ reads
\begin{equation*}
  \hat{s}(x,k) = 1-\hat{\pi}(x,k),
\end{equation*}
where $\hat{\pi}(x,k)$ denotes the probability estimated by the black-box classifier for an observation with features $x$ to have label $k$.

For a point in the calibration set with features $X_i$ and label $Y_i$, we compute the nonconformity score as

\begin{align*}
  \hat{s}(X_i,Y_i) = 1-\hat{\pi}(X_i,Y_i),
\end{align*}

where $\hat{\pi}(X_i,Y_i)$ denotes the probability estimated by the black-box classifier for the observation $i$ to have the true label $Y_i$. These scores are then utilized to calibrate a prediction set for the test point $(X_{n+1}, Y_{n+1})$.

For a desired significance level $\alpha \in (0,1)$, we calculate the $\lceil (1+n)\cdot(1-\alpha) \rceil$-th smallest value in $\{\hat{s}(X_i,Y_i)\}_{i \in D^{\mathrm{cal}}}$. This value is used to define a threshold $\hat{\tau}$:

\begin{equation*}
  \hat{\tau} := \lceil (1+n)\cdot(1-\alpha) \rceil \text{-th smallest value of the scores}.
\end{equation*}

Then, the prediction set for $Y_{n+1}$ is identified as the set of all labels $k \in [K]$ for which $\hat{\pi}(x,k)$ is larger than $\hat{\tau}$; i.e.,
\begin{equation}\label{eq:prediction_set}
  \hat{C}(X_{n+1}) := \{ k \in [K] : \hat{\pi}(x,k) \geq 1 - \hat{\tau} \}.
\end{equation}

This procedure guarantees that $\hat{C}(X_{n+1})$ has **marginal coverage** at level $1 - \alpha$.

### Implementation aspects

Function for data generation for a classification problem.

```{r}
## Simulate data for a classification problem
generate_data_class <- function(n, p, K, seed = NULL) {
  # n: Total number of observations
  # p: Number of features
  # K: Number of classes
  # seed: Optional seed for reproducibility
  
  if (!is.null(seed)) {
    set.seed(seed)
  }
  
  ## Generate features X
  X <- matrix(rnorm(n * p, mean = 0, sd = 1), nrow = n, ncol = p)
  X[, 1] <- sample(c(-1, 1), n, replace = TRUE, prob = c(1 / K, (K - 1) / K))
  X[, 2] <- sample(c(-1, 1), n, replace = TRUE, prob = c(1 / 4, 3 / 4))
  X[, 3] <- sample(c(-1, 1), n, replace = TRUE, prob = c(1 / 2, 1 / 2))
  X[, 4] <- sample(1:K, n, replace = TRUE)
  
  ## Compute class probabilities according to X
  prob_y <- matrix(0, nrow = nrow(X), ncol = K)
  
  # Split the feature space in 4 regions
  leaf_1 <- which(X[, 1] <= 0)
  leaf_2 <- which(X[, 1] > 0 & X[, 2] <= 0 & X[, 3] <= 0)
  leaf_3 <- which(X[, 1] > 0 & X[, 2] <= 0 & X[, 3] > 0)
  leaf_4 <- which(X[, 1] > 0 & X[, 2] > 0)
  
  # Leaf 1: Uniform over all classes
  prob_y[leaf_1, ] <- 1 / K
  
  # Leaf 2: Uniform over first half of classes
  K_half <- ceiling(K / 2)
  prob_y[leaf_2, 1:K_half] <- 2 / K
  prob_y[leaf_2, (K_half + 1):K] <- 0
  
  # Leaf 3: Uniform over second half of classes
  prob_y[leaf_3, 1:K_half] <- 0
  prob_y[leaf_3, (K_half + 1):K] <- 2 / K
  
  # Leaf 4: 90% probability to the class specified by the 4th feature
  if (length(leaf_4) > 0) {
    X4 <- X[leaf_4, 4]
    for (k in 1:K) {
      idx <- leaf_4[X4 == k]
      if (length(idx) > 0) {
        prob_y[idx, k]  <- 0.9
        prob_y[idx, -k] <- (1 - 0.9) / (K - 1)
        
      }
    }
  }
  prob_y <- prob_y / rowSums(prob_y) # Normalize probabilities
  
  ## Sample labels Y
  Y <- apply(prob_y, 1, function(probs) {
    sample(1:K, size = 1, prob = probs)
  })

  return(list(X = X, Y = Y, prob_y = prob_y))
}

```

We consider a simple classification example, where each label can take value in 1,...,K.

```{r}
# Generate data
n <- 2000
p <- 8
K <- 10
data <- generate_data_class(n = n, p = p, K = K, seed = 2025)
X_data <- data$X
Y_data <- data$Y
```

We generate a test set, which we will use to evaluate the predictions made by the conformal prediction algorithm.

```{r}
# Generate independent test data
n_test <- 1000
data_test <- generate_data_class(n = n_test, p = p, K = K, seed = 2025+n)
X_test <- data_test$X           
Y_test <- data_test$Y
```


#### Split conformal for classification

```{r}
#' Compute conformal prediction interval
sc_classification <- function(X, Y, X_test, alpha, seed = 2030, allow_empty=TRUE){
  
  # Inputs:
  # X      : Matrix of explanatory variables
  # Y      : Vector of response labels
  # X_test : Matrix of explanatory variables for test data
  # alpha  : target significance level
  # seed   : Seed for reproducibility
  
  set.seed(seed)
  
  # Split the data into training and calibration sets
  n <- nrow(X)
  p <- ncol(X)
  train_index <- sample(1:n, size = floor(0.5 * n))
  X_train <- X[train_index, ]
  Y_train <- Y[train_index]
  
  X_calib <- X[-train_index, ]
  Y_calib <- Y[-train_index]
  n.calib <- dim(X_calib)[1]
  
  data_train <- data.frame(cbind(Y_train, X_train))
  names(data_train) <- c("Y", paste0("V", seq_len(ncol(X_train))))
  data_train$Y <- as.factor(data_train$Y)
  
  # Fit any black box classifier (in this case it is a random forest)
  black_box <- randomForest(Y ~ ., ntree=500, mtry=p, data = data_train)
  
  # Estimate class probabilities for calibration points
  pi_hat_calib <- predict(black_box, newdata = X_calib, type = "prob")

  # Break ties at random
  pi_hat_calib <- pi_hat_calib + 1e-9 * runif(length(pi_hat_calib), min = -1, max = 1)
  pi_hat_calib <- pi_hat_calib / rowSums(pi_hat_calib) # Normalize rows to sum to 1
  
  # Extract the predicted probability corresponding to the true label
  # for each observation in the calibration dataset
  pi_y_calib <- sapply(1:length(Y_calib), function(i) pi_hat_calib[i, Y_calib[i]])
  
  # Evaluate the homogeneous scores
  scores <- 1 - pi_y_calib
  
  # Compute threshold (what we called tau_hat)
  ncs.calib.sort <- c(sort(scores), Inf)
  ncs.quantile <- ncs.calib.sort[ceiling((n.calib + 1)*(1-alpha))]
  
  # Estimate probabilities for the test points
  pi_hat_test <- predict(black_box, newdata = X_test, type = "prob")
  
  # Break ties at random
  pi_hat_test <- pi_hat_test + 1e-9 * runif(length(pi_hat_test), min = -1, max = 1)
  pi_hat_test <- pi_hat_test / rowSums(pi_hat_test) # Normalize rows to sum to 1
  
  # Build the prediction sets
  n.test <- dim(X_test)[1]
  S_hat <- vector("list", n.test)
  for (i in 1:n.test) {
    S_hat[[i]] <- which(pi_hat_test[i, ] >= 1 - ncs.quantile)
    if (!allow_empty && length(S_hat[[i]]) == 0) {
      S_hat[[i]] <- which.max(pi_hat_test[i, ])
    }
  }
  
  return(S_hat)
  
}

```


Function for evaluating the conformal prediction sets.

```{r}
## Function for evaluating predictions
evaluate_predictions <- function(S, X, Y, verbose = TRUE) {
  # S: List of length n, where each element is a vector of prediction sets
  # X: n x p data matrix of explanatory variables
  # Y: n-length vector of true response labels
  # verbose: Whether to print the metrics
  
  n <- length(Y)
  
  # Compute marginal coverage
  marginal_coverage <- mean(sapply(1:n, function(i) Y[i] %in% S[[i]]))
  
  # Compute average size of prediction sets
  size <- mean(sapply(S, length))
  
  # Compute average size of prediction sets that contain the true label
  idx_cover <- which(sapply(1:n, function(i) Y[i] %in% S[[i]]))
  size_cover <- if (length(idx_cover) > 0) {
    mean(sapply(S[idx_cover], length))
  } else {
    NA
  }
  
  # Print summary
  if (verbose) {
    cat(sprintf("Marginal coverage       : %2.3f%%\n", marginal_coverage * 100))
    cat(sprintf("Average size            : %2.3f\n", size))
    cat(sprintf("Average size | cover    : %2.3f\n", ifelse(is.na(size_cover), NA, size_cover)))
  }
  
  # Return results as a list
  return(list(
    marginal_coverage = marginal_coverage,
    average_size = size,
    average_size_cover = size_cover
  ))
}

```

Usage

```{r}
# Desired coverage level (1-alpha)
alpha <- 0.1

# Apply split conformal classification
S <- sc_classification(X_data, Y_data, X_test, alpha)

# Evaluate prediction sets
metrics <- evaluate_predictions(S, X_test, Y_test)

```

